A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-24 02-ceu.tex
Consertos pequenos na Seção 3

Listing~\ref{lst.syntax} shows a compact reference of~\CEU:

@@

In \CEU, queue-based processing of events applies only to external input
events, i.e., events submitted to the program by the environment.
%
Internal events, which are events generated internally by the program via
\code{emit} statements, are processed in a stack-based manner.
%
Internal events provide a fine-grained execution control, and, because of their
stack-based processing, can be used to implement a limited form of subroutines,
as illustrated in Listing~\ref{lst.sub}:

>>>>

Listing~\ref{lst.syntax} shows a compact reference of~\CEU.

@@

In \CEU, queue-based processing of events applies only to external input
events, i.e., events submitted to the program by the environment.
%
Internal events, which are events generated internally by the program via
\code{emit} statements, are processed in a stack-based manner.
%
Internal events provide a fine-grained execution control, and, because of their
stack-based processing, can be used to implement a limited form of subroutines,
as illustrated in Listing~\ref{lst.sub}.

### 2018-02-24 03-semantics.tex
Consertos pequenos na Seção 3

In this section, we introduce a reduced syntax of \CEU and propose an
operational semantics to formally describe the behavior of programs.
We describe a small synchronous kernel highlighting the peculiarities of \CEU,
in particular, the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language,
leaving out side-effects and system calls (which behave like in conventional
imperative languages).

@@

The core of our semantics describes how a program reacts to a single external
input event, i.e., starting from an input event, how the program behaves and
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules~\cite{Plotkin-G-D-1981}, which
are designed in such a way that at most one transition is possible at any
time, resulting in deterministic reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are non-negative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both being possibly the
empty event~$\nil$).
$\E$ is a set of primitive events.

We will refer to the triples on the left-hand and right-hand sides of
symbol~$\to$ as \emph{descriptions} (denoted~$\delta$).  The triple on the
left-hand side of symbol~$\to$ is called the \emph{input description}, and
the triple on its right-hand side is called the \emph{output description}.

@@

The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions which only
apply to the program as a whole, and are the only to manipulate the stack
level:
\begin{align*}
  &\AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[\JOT]
  %%
  &\AxiomC{$n>0$}
  \AxiomC{$p=\ceu{\Nop,\Break}\vee\isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{align*}

@@

The rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions make sure that their
left branches always transition before their right branches:
%%
\begin{gather*}
  \hskip-.9em
  \<\ceu{p\And{q}},n,\nil>
  \nst\<\ceu{p\AtAnd(\CanRun(n);q)},n,\nil>
  \Rtag{and-expd}\\[\JOT]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p'\AtAnd{q}},n,e>$}
  \DisplayProof
  \Rtag{and-adv1}\\[\JOT]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
  \BinaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p\AtAnd{q'}},n,e>$}
  \DisplayProof
  \Rtag{and-adv2}
\end{gather*}

@@

  More formally, $\pot(p,e)=\pot'(\bcast(p,e))$, where~$\pot'$ is an
  auxiliary function that counts the number of reachable~$\ceu{\EmitInt}$
  expressions in the program resulting from the broadcast of event~$e$
  to~$p$.

@@

An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
transitions~$\trans$ in general.  Such descriptions cannot be advanced
by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
nor~$\outpop$, as the second coordinate of its rank is~0, which
implies~$e=\nil$ and~$n=0$.

The next two lemmas establish that a single application of~$\out$ or~$\nst$
either preserves or decreases the rank of the input description.  All rank
comparisons assume lexicographical order, i.e., if~$\rank(\delta)=\<i,j>$
and~$\rank(\delta')=\<i',j'>$ then~$\rank(\delta)>\rank(\delta')$ iff~$i>i'$
or~$i=i$ and~$j>j'$.
%%
The proof of Lemma~\ref{lem.x.rank-out} follows directly from~\R{push}
and~\R{pop} and from Definitions~\ref{def.x.pot} and~\ref{def.x.rank}.  The
proof of Lemma~\ref{lem.x.rank-nst}, however, is by induction on the
structure of~$\nst$ derivations.  Since Lemma~\ref{lem.x.rank-nst} is of
fundamental importance for the termination result, we detail the proof of
one of its cases.

@@

  We proceed to prove the case where~$p=\ceu{{p_1}\AtLoop{p_2}}$, which
  itself consists of three subcases.
  \begin{case}
    $p_1=\ceu{\Nop}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-nop}.  i.e., its conclusion is
    obtained by an application of this rule.  Hence~$p'=p_2$ and~$e'=\nil$.
    Thus
    \begin{align*}
        \rank(\delta)=\<\pot'(p_1)+\pot'(p_2),n>
                     \ge\<pot'(p_2),n>=\rank(\delta')\,.
    \end{align*}
  \end{case}
  \begin{case}
    $p_1=\ceu{\Break}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-brk}.
    Hence~$p'=p_1$ and~$e'=\nil$.  Thus
    $\rank(\delta)=\rank(\delta')=\<0,n>$.
    %%
    % \[
    %   \rank(\delta)=\rank(\delta')=\<0,n>\,.
    % \]
    %%
  \end{case}
  \begin{case}
    $p_1\ne\ceu{\Nop},\ceu{\Break}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-adv}.  Hence there is a
    derivation~$\pi'$ such that
    \begin{equation}
      \label{lem.x.rank-nst-eq1}
      \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
    \end{equation}
    for some~$p_1'$ and~$e_1'$.  Thus~$p'=\ceu{p_1'\AtLoop{\,p_2}}$
    and~$e'=e_1'$.  There are two subcases.
    \begin{subcase}
      \label{lem.x.rank-nst-case1}
      $\pot'(p)=\pot'(p_1)$.
      %%
      Then, by Definition~\ref{def.pot}, every execution path of~$p_1$
      contains a (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
      A single~$\nst$ cannot terminate the loop, since~$p_1\ne\ceu{\Break}$,
      nor can it consume an~$\ceu{\AwaitExt}$, which means that all
      execution paths in~$p_1'$ still contain a~$\ceu{\Break}$
      or~$\ceu{\AwaitExt}$.  Hence~$\pot'(p')=\pot'(p_1')$.

      There are two further subcases depending on event~$e'$.
      \begin{subsubcase}
        $e'=\nil$.
        %%
        Then $\rank(\delta)=\<pot'(p_1),n>$ and
        $\rank(\delta')=\<pot'(p_1'),n>$.  From~\eqref{lem.x.rank-nst-eq1},
        by the induction hypothesis, $\pot'(p_1)\ge\pot(p_1')$.  Thus
        $\rank(\delta)\ge\rank(\delta')$.
        %%
        % \[
        %   \rank(\delta)\ge\rank(\delta')\,.
        % \]
        %%
      \end{subsubcase}
      \begin{subsubcase}
        $e'\ne\nil$.
        %%
        Then~$\pi'$ contains one application of~\R{emit-int} which consumes
        one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and which implies
        $\pot'(p_1)>\pot'(p_1')$.  Thus
        \[
          \hskip1em\rank(\delta)\!=\!\<\pot'(p_1),n>
          >\<\pot'(p_1'),n+1>\!=\!\rank(\delta')\,.
        \]
        %%
        % \begin{align*}
        %   \rank(\delta)&=\<\pot'(p_1),n>\\
        %                &>\<\pot'(p_1'),n+1>=\rank(\delta')\,.
        % \end{align*}
        %%
      \end{subsubcase}
    \end{subcase}
    \begin{subsubcase}
      $\pot'(p)=\pot'(p_1)+\pot'(p_2)$.
      %%
      Then some execution path in~$p_1$ does not contain a
      (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
      Since~$p_1\ne\ceu{\Nop}$, a single~$\nst$ cannot restart the loop,
      which means that~$p_1'$ still contain some execution path in which
      a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ does not occur.
      Hence~$\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
      similar to that of Case~\ref{lem.x.rank-nst-case1}.\qedhere
    \end{subsubcase}
  \end{case}
\end{proof}

@@

We now state and prove the main result of this section,
Theorem~\ref{thm.x.term}, viz., the termination theorem for~$\trans[*]$.
The idea of the proof is that a series of~$\nst$ or~$\out$ transitions
eventually decrease the rank of the current description until an irreducible
description is reached.  This irreducible description is the final result of
the reaction.
%%
\begin{theorem}[Termination]\label{thm.x.term}\strut\\
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$\rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.
  \begin{basis}
    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
    applied).  There are two possibilities:  either~$\delta$ is
    nested-irreducible or it is not.  In the first case, the theorem is
    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
    nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
    Theorem~\ref{thm.x.rank-nst-*},
    $\<i,j>=\<0,0>\ge\rank(\delta')$,
    %%
    % \[
    %   \<i,j>=\<0,0>\ge\rank(\delta')\,,
    % \]
    %%
    which implies~$\rank(\delta')=\<0,0>$.
  \end{basis}
  \begin{induction}
    Let~$\<i,j>\ne\<0,0>$.
    %%
    There are two subcases depending on whether or not $\delta$~is
    nested-irreducible.
    \begin{case}
      \label{thm.x.term.Hnst}
      $\delta$~is nested-irreducible.
      %%
      If~$j=0$ then, by Definition~\ref{def.x.H}, $\delta_\#$, and
      thus~$\delta\trans[0]\delta_\#$.  Suppose that is not the case, i.e.,
      that $j>0$.  Then there are two subcases.
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nonnil}
        $e\ne\nil$.
        %%
        Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
        are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
        $\delta\outpush\delta_1'\nst[*]\delta'_\Hnst$.  Thus, by
        Lemma~\ref{lem.x.rank-out} and by Theorem~\ref{thm.x.rank-nst-*},
        \begin{align*}
          \rank(\delta)=\rank(\delta_1')=\<i,j>
                                        \ge\rank(\delta')=\<i',j'>\,.
        \end{align*}
        If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof is
        similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil} below.
        Otherwise, if~$e'\ne\nil$ then $i>i'$, since an~$\ceu{\EmitInt(e')}$
        was consumed by the nested transitions.  Thus,
        $\rank(\delta)>\rank(\delta')$.  By the induction hypothesis,
        $\delta'\trans[*]\delta''_\#$, for some~$\delta''_\#$.  Therefore,
        $\delta\trans[*]\delta''_\#$.
      \end{subcase}
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nil}
        $e=\nil$.
        %%
        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta''$.  By
        item~\eqref{lem.rank-out-pop} of Lemma~\ref{lem.rank-out},
        $\rank(\delta)>\rank(\delta')$.  Hence, by the induction hypothesis,
        there is a~$\delta''_\#$ such that~$\delta'\trans[*]\delta''_\#$.
        Therefore, $\delta\trans[*]\delta''_\#$.
      \end{subcase}
    \end{case}
    \begin{case}
      $\delta$~is not nested-irreducible.
      %%
      Then~$e=\nil$ and, by Theorems~\ref{thm.x.rank-nst-*}
      and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
      that~$\delta\nst[*]\delta'_\Hnst$
      with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
      is similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
    \end{case}
  \end{induction}
\end{proof}

@@

As \CEU has no mechanism for heap allocation, unbounded iteration, or
general recursion, the maximum memory usage of a given \CEU program is
determined solely by the length of its code, the number of variables it
uses, and the size of the event stack that it requires to run.  The code
length and the number of variables used are easily determined by code
inspection.  The maximum size of the event stack during a reaction of
program~$p$ to external event~$e$ corresponds to~$\pot(p,e)$, i.e., to the
maximum number of internal events that~$p$ may emit in a reaction to~$e$.
If~$p$ may react to external events~$e_1$, \dots,~$e_n$ then, in the worst
case, its event stack will need to
store~$\max\{\pot(p,e_1),\dots,\pot(p,e_n)\}$ events.

>>>>

In this section, we introduce a reduced syntax for \CEU and propose an
operational semantics to formally describe the behavior of its programs.
We describe a small synchronous kernel highlighting the peculiarities of \CEU,
in particular, the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language,
leaving out side-effects and system calls (which behave like in conventional
imperative languages).

@@

The core of our semantics describes how a program reacts to a single external
input event, i.e., starting from an input event, how the program behaves and
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules~\cite{Plotkin-G-D-1981} which
are designed in such a way that at most one transition is possible at any
time, resulting in deterministic reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are nonnegative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both possibly being the
empty event~$\nil$).
$\E$ is a set of primitive events.

We will refer to the triples on the left-hand and right-hand sides of
symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The triple on the
left-hand side of symbol~$\trans$ is called the \emph{input description}, and
the triple on its right-hand side is called the \emph{output description}.

@@

The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions that
apply to the program as a whole.  These are the only rules that manipulate
the stack level.
\begin{align*}
  &\AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[\JOT]
  %%
  &\AxiomC{$n>0$}
  \AxiomC{$p=\ceu{\Nop,\Break}\vee\isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{align*}

@@

The rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions ensure that the
left-hand side of compositions always transition before their right-hand
side:
%%
\begin{gather*}
  \hskip-.9em
  \<\ceu{p\And{q}},n,\nil>
  \nst\<\ceu{p\AtAnd(\CanRun(n);q)},n,\nil>
  \Rtag{and-expd}\\[\JOT]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p'\AtAnd{q}},n,e>$}
  \DisplayProof
  \Rtag{and-adv1}\\[\JOT]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
  \BinaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p\AtAnd{q'}},n,e>$}
  \DisplayProof
  \Rtag{and-adv2}
\end{gather*}

@@

  More formally, $\pot(p,e)=\pot'(\bcast(p,e))$ where~$\pot'$ is an
  auxiliary function that counts the number of reachable~$\ceu{\EmitInt}$
  expressions in the program resulting from the broadcast of event~$e$
  to~$p$.

@@

An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
transitions~$\trans$ in general.  Such description cannot be advanced
by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
nor~$\outpop$, as the second coordinate of its rank is~0, which
implies~$e=\nil$ and~$n=0$.

The next two lemmas establish that a single application of~$\out$ or~$\nst$
either preserves or decreases the rank of the input description.  All rank
comparisons assume lexicographic order, i.e., if~$\rank(\delta)=\<i,j>$
and~$\rank(\delta')=\<i',j'>$ then~$\rank(\delta)>\rank(\delta')$ iff~$i>i'$
or~$i=i$ and~$j>j'$.
%%
The proof of Lemma~\ref{lem.x.rank-out} follows directly from~\R{push}
and~\R{pop} and from Definitions~\ref{def.x.pot} and~\ref{def.x.rank}.  The
proof of Lemma~\ref{lem.x.rank-nst}, however, is by induction on the
structure of~$\nst$ derivations.  Since Lemma~\ref{lem.x.rank-nst} is of
fundamental importance for the termination result, we detail the proof of
one of its cases.

@@

  We proceed to prove the case where~$p=\ceu{{p_1}\AtLoop{p_2}}$, which
  itself consists of three subcases.
  \begin{case}
    $p_1=\ceu{\Nop}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-nop}.  i.e., its conclusion is
    obtained by an application of this rule.  Hence~$p'=p_2$ and~$e'=\nil$.
    Thus
    \begin{align*}
        \rank(\delta)=\<\pot'(p_1)+\pot'(p_2),n>
                     \ge\<pot'(p_2),n>=\rank(\delta')\,.
    \end{align*}
  \end{case}
  \begin{case}
    $p_1=\ceu{\Break}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-brk}.
    Hence~$p'=p_1$ and~$e'=\nil$.  Thus
    $\rank(\delta)=\rank(\delta')=\<0,n>$.
    %%
    % \[
    %   \rank(\delta)=\rank(\delta')=\<0,n>\,.
    % \]
    %%
  \end{case}
  \begin{case}
    $p_1\ne\ceu{\Nop},\ceu{\Break}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-adv}.  Hence there is a
    derivation~$\pi'$ such that
    \begin{equation}
      \label{lem.x.rank-nst-eq1}
      \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
    \end{equation}
    for some~$p_1'$ and~$e_1'$.  Thus~$p'=\ceu{p_1'\AtLoop{\,p_2}}$
    and~$e'=e_1'$.

    There are two subcases.
    \begin{subcase}
      \label{lem.x.rank-nst-case1}
      $\pot'(p)=\pot'(p_1)$.
      %%
      Then, by Definition~\ref{def.pot}, every execution path of~$p_1$
      contains a (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
      A single~$\nst$ cannot terminate the loop, since~$p_1\ne\ceu{\Break}$,
      nor it can consume an~$\ceu{\AwaitExt}$, which means that all
      execution paths in~$p_1'$ still contain a~$\ceu{\Break}$
      or~$\ceu{\AwaitExt}$.  Hence~$\pot'(p')=\pot'(p_1')$.

      There are two subcases.
      \begin{subsubcase}
        $e'=\nil$.
        %%
        Then $\rank(\delta)=\<pot'(p_1),n>$ and
        $\rank(\delta')=\<pot'(p_1'),n>$.  From~\eqref{lem.x.rank-nst-eq1},
        by the induction hypothesis, $\pot'(p_1)\ge\pot(p_1')$.  Thus
        $\rank(\delta)\ge\rank(\delta')$.
        %%
        % \[
        %   \rank(\delta)\ge\rank(\delta')\,.
        % \]
        %%
      \end{subsubcase}
      \begin{subsubcase}
        $e'\ne\nil$.
        %%
        Then~$\pi'$ contains one application of~\R{emit-int} which consumes
        one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and which implies
        $\pot'(p_1)>\pot'(p_1')$.  Thus
        \[
          \hskip1em\rank(\delta)\!=\!\<\pot'(p_1),n>
          >\<\pot'(p_1'),n+1>\!=\!\rank(\delta')\,.
        \]
        %%
        % \begin{align*}
        %   \rank(\delta)&=\<\pot'(p_1),n>\\
        %                &>\<\pot'(p_1'),n+1>=\rank(\delta')\,.
        % \end{align*}
        %%
      \end{subsubcase}
    \end{subcase}
    \begin{subsubcase}
      $\pot'(p)=\pot'(p_1)+\pot'(p_2)$.
      %%
      Then some execution path in~$p_1$ does not contain a
      (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
      Since~$p_1\ne\ceu{\Nop}$, a single~$\nst$ cannot restart the loop,
      which means that~$p_1'$ still contain some execution path in which
      a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ does not occur.
      Hence~$\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
      similar to that of Case~\ref{lem.x.rank-nst-case1}.\qedhere
    \end{subsubcase}
  \end{case}
\end{proof}

@@

We now state and prove the main result of this section,
Theorem~\ref{thm.x.term}, viz., the termination theorem for~$\trans[*]$.
The idea of the proof is that a series of~$\nst$ or~$\out$ transitions
eventually decrease the rank of the current description until an irreducible
description is reached.  This irreducible description is the final result of
the reaction.
%%
\begin{theorem}[Termination]\label{thm.x.term}\strut\\
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$\rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.
  \begin{basis}
    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
    applied).  There are two possibilities:  either~$\delta$ is
    nested-irreducible or it is not.  In the first case, the theorem is
    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
    nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
    Theorem~\ref{thm.x.rank-nst-*},
    $\<i,j>=\<0,0>\ge\rank(\delta')$,
    %%
    % \[
    %   \<i,j>=\<0,0>\ge\rank(\delta')\,,
    % \]
    %%
    which implies~$\rank(\delta')=\<0,0>$.
  \end{basis}
  \begin{induction}
    Let~$\<i,j>\ne\<0,0>$.
    %%
    There are two subcases depending on whether or not $\delta$~is
    nested-irreducible.
    \begin{case}
      \label{thm.x.term.Hnst}
      $\delta$~is nested-irreducible.
      %%
      If~$j=0$ then, by Definition~\ref{def.x.H}, $\delta_\#$, and
      thus~$\delta\trans[0]\delta_\#$.  Suppose that is not the case, i.e.,
      that $j>0$.  Then there are two subcases.
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nonnil}
        $e\ne\nil$.
        %%
        Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
        are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
        $\delta\outpush\delta_1'\nst[*]\delta'_\Hnst$.  Thus, by
        Lemma~\ref{lem.x.rank-out} and by Theorem~\ref{thm.x.rank-nst-*},
        \begin{align*}
          \rank(\delta)=\rank(\delta_1')=\<i,j>
                                        \ge\rank(\delta')=\<i',j'>\,.
        \end{align*}
        If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof is
        similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil} below.
        Otherwise, if~$e'\ne\nil$ then $i>i'$, since an~$\ceu{\EmitInt(e')}$
        was consumed by the nested transitions.  Thus,
        $\rank(\delta)>\rank(\delta')$.  By the induction hypothesis,
        $\delta'\trans[*]\delta''_\#$, for some~$\delta''_\#$.  Therefore,
        $\delta\trans[*]\delta''_\#$.
      \end{subcase}
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nil}
        $e=\nil$.
        %%
        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta'$.  By
        item~\eqref{lem.rank-out-pop} of Lemma~\ref{lem.rank-out},
        $\rank(\delta)>\rank(\delta')$.  Hence, by the induction hypothesis,
        there is a~$\delta''_\#$ such that~$\delta'\trans[*]\delta''_\#$.
        Therefore, $\delta\trans[*]\delta''_\#$.
      \end{subcase}
    \end{case}
    \begin{case}
      $\delta$~is not nested-irreducible.
      %%
      Then~$e=\nil$ and, by Theorems~\ref{thm.x.rank-nst-*}
      and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
      that~$\delta\nst[*]\delta'_\Hnst$
      with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
      is similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
    \end{case}
  \end{induction}
\end{proof}

@@

As \CEU has no mechanism for heap allocation, unbounded iteration, or
general recursion, the maximum memory usage of a given \CEU program is
determined solely by the length of its code, the number of variables it
uses, and the size of the event stack that it requires to run.  The code
length and the number of variables used are easily determined by code
inspection.  The maximum size of the event stack during a reaction of
program~$p$ to external event~$e$ corresponds to~$\pot(p,e)$, i.e., to the
maximum number of internal events that~$p$ may emit in reaction to~$e$.
If~$p$ may react to external events~$e_1$, \dots,~$e_n$ then, in the worst
case, its event stack will need to
store~$\max\{\pot(p,e_1),\dots,\pot(p,e_n)\}$ events.
